Embracing the Laws of Physics: \\ Reversible Models of Computation

### 2017-11-28 survey/survey.tex
plan for survey

@@

In our current age, both Laudauer's Princple and recent investigations
of computational models well-suited to quantum mechanics, push us to
revisit our current models of computation. Information, on top of
having a physical manifestation, appears to be \emph{conserved},
along with other quantities like energy, mass and momentum.

@@

\section{Models of Computation}
%%%

@@

These irreversible abstractions chosen as the basis of our
models of computing are at odds with the underlying
reversible physical reality. The information theoretic underpinnings
of physics is best understood through \emph{Landauer's
principle}~\cite{Landauer:1961}. As mentionned previously in the
introduction, Landauer, through his analysis of
Maxwell's Demon, established the idea that \emph{information is a
physical quantity} in much the sense that one would think of an
electron as a physical quantity. More precisely, his principle implies
that a certain minimum amount of
thermodynamic work has to be carried out to erase one bit of
information. Experimental verification of the Landauer principle
happened in recent years~\cite{berut2012experimental}. If information
has a physical significance, then it is subject to conservation laws
just as mass and energy are.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conservation of Information}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Reversibility}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusions}

>>>>

% * Reversibility intro / motivation as we  have now more or less

% * Thesis statement: programs are reversible deformations on data /
%    spaces, almost necessarily by definition

% * Focus now is on data

% * If data is plain finite sets, programs become permutations (deformations on finite sets).

% * If data is structured trees, we get Pi.

% * Explain Pi with examples.

% * If data is itself deformations-on-finite-sets, then the
%   deformations between them become something quite
%   interesting. Explain Pi level 2 with examples.

% * Conclude with thoughts regarding other kinds of data that can be
%   plugged in into that story.

@@

In our current age, both Landauer's Principle and recent investigations
of computational models well-suited to quantum mechanics, push us to
revisit our current models of computation. Information, on top of
having a physical manifestation, appears to be \emph{conserved},
along with other quantities like energy, mass and momentum.

@@

@@

These irreversible abstractions chosen as the basis of our
models of computing are at odds with the underlying
reversible physical reality. The information theoretic underpinnings
of physics is best understood through \emph{Landauer's
principle}~\cite{Landauer:1961}. As mentioned previously in the
introduction, Landauer, through his analysis of
Maxwell's Demon, established the idea that \emph{information is a
physical quantity} in much the sense that one would think of an
electron as a physical quantity. More precisely, his principle implies
that a certain minimum amount of
thermodynamic work has to be carried out to erase one bit of
information. Experimental verification of the Landauer principle
happened in recent years~\cite{berut2012experimental}. If information
has a physical significance, then it is subject to conservation laws
just as mass and energy are.

@@

@@

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Programs as Reversible Deformations}

Starting from a physical perspective, programs naturally emerge as
reversible transformations on data. In some sense, there is nothing
more to say about programs per se. The focus is on the data. Depending
on the space we use to represent the data, the corresponding notion of
deformation on these spaces will define the appropriate notion of
programs. In the following three sections, we consider three natural
classes of data and explore the corresponding notion of reversible
programs.

In the quantum case, this is clear. Say something about topological
quantum computing.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Data I: Finite Sets}

Most programming languages provide a rich collection of primitive
data like booleans, characters, strings, and numbers that are
naturally modeled as sets. Infinite sets embody a notion of recursion
which is subtle and is discussed briefly in the concluding
section. For finite sets, a reversible process between them is simply
a permutation. In other words, reversible programming with finite sets
is all about writing permutations.

We therefore start with a small language for writing permutations as a
product of transpositions:

\begin{verbatim}
p ::= id | swap i j | p;p
\end{verbatim}

Permutations on finite sets of size $n$ can be inductively defined as
follows:
\begin{verbatim}
perms(0)   = { id }
perms(n+1) = { (swap 0 i).p | p <- perms(n); i <- [1..n] }  
\end{verbatim}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Data II: Structured Finite Types}

Pi level 1

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Data III: Reversible Programs between Structured Finite Types}

Pi level 2

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusion}

Discuss recursion, or add a section of infinite sets???
